---
title: "Index Graphs"
author: "Mike Finnegan"
date: "2/4/2018"
output: pdf_document
---

**Dissimilarity Index:** $$D_t=\frac{1}{2}\sum_{i=1}^{n}|\frac{b_{it}}{B_{t}}-\frac{w_{it}}{W_{t}}|$$

Interpretable as the percent of black students who would need to be reassigned to a different school for perfect integration to be achieved given the district’s overall racial composition. Ranges from 0 to 1, with 0 denoting perfect integration and 1 denoting
complete segregation


**Exposure Index:** $$E_t=\frac{1}{B_{t}}\sum_{i=1}^{n}b_{it}*\frac{w_{it}}{t_{it}}$$

Interpretable as the percent of white students in the average black student’s school. For a given county/district, it ranges from 0 to the percent of white students in the county/district as a whole.



# Publics and Privates

## Dissimilarity

```{r pub and pvt: dissimilarity, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
pub.index <- read.csv("~/Documents/Github/Schools/pub.index.csv")
pvt.index <- read.csv("~/Documents/Github/Schools/pvt.index.csv")

df.schools <- rbind(pub.index, pvt.index)

################################
# Dissimilarity Index: County
################################

df.schools <- df.schools %>% 
  group_by(COUNTY, YEAR) %>%
  mutate(county.total=sum(NUMSTUDS),
         black.county.total=sum(BLACK),
         white.county.total=sum(WHITE))

df.schools$pct.black.county.lvl <- df.schools$BLACK/df.schools$black.county.total
df.schools$pct.white.county.lvl <- df.schools$WHITE/df.schools$white.county.total
df.schools$pct.diff.county.lvl <- abs(df.schools$pct.black.county.lvl-df.schools$pct.white.county.lvl)
df.schools[is.na(df.schools)] <- 0

dissimilarity.county.level <- df.schools %>% 
  group_by(COUNTY, YEAR, DISMISSED) %>%
  summarise(county.index=0.5*sum(pct.diff.county.lvl))

avg.diss.county.both <- dissimilarity.county.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.county.index = mean(county.index))
avg.diss.county.both$DISMISSED <- as.factor(avg.diss.county.both$DISMISSED)
avg.diss.county.both$YEAR <- as.factor(avg.diss.county.both$YEAR)


ggplot(avg.diss.county.both, aes(x=YEAR, average.county.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Dissimilarity Index")+
  ggtitle("Average Dissimilarity Index by Dismissal Status on County
          Level Including Publics and Privates")

################################
# Dissimilarity Index: District
################################

df.schools <- df.schools %>% 
  group_by(DISTRICT, YEAR) %>%
  mutate(district.total=sum(NUMSTUDS),
         black.district.total=sum(BLACK),
         white.district.total=sum(WHITE))

df.schools$pct.black.district.lvl <- df.schools$BLACK/df.schools$black.district.total
df.schools$pct.white.district.lvl <- df.schools$WHITE/df.schools$white.district.total
df.schools$pct.diff.district.lvl <- abs(df.schools$pct.black.district.lvl-df.schools$pct.white.district.lvl)
df.schools[is.na(df.schools)] <- 0

dissimilarity.district.level <- df.schools %>% 
  group_by(DISTRICT, YEAR, DISMISSED) %>%
  summarise(district.index=0.5*sum(pct.diff.district.lvl))

avg.diss.district.both <- dissimilarity.district.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.district.index = mean(district.index))
avg.diss.district.both$DISMISSED <- as.factor(avg.diss.district.both$DISMISSED)
avg.diss.district.both$YEAR <- as.factor(avg.diss.district.both$YEAR)

ggplot(avg.diss.district.both, aes(x=YEAR, average.district.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Dissimilarity Index")+
  ggtitle("Average Dissimilarity Index by Dismissal Status on District
          Level Including Publics and Privates")


```

## Exposure
```{r pub and pvt: exposure, echo=FALSE}

################################
# Exposure Index: County
################################

df.schools$pct.to.sum <- (df.schools$WHITE/df.schools$NUMSTUDS)*df.schools$BLACK

exp.county.level <- df.schools %>% 
  group_by(COUNTY, YEAR, DISMISSED) %>%
  summarise(county.index=sum((WHITE/NUMSTUDS)*BLACK),
            black.county.total=sum(BLACK))
exp.county.level[is.na(exp.county.level)] <- 0
exp.county.level$county.index <- exp.county.level$county.index/exp.county.level$black.county.total
exp.county.level[is.na(exp.county.level)] <- 0


avg.exp.county.both <- exp.county.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.county.index = mean(county.index))
avg.exp.county.both$DISMISSED <- as.factor(avg.exp.county.both$DISMISSED)
avg.exp.county.both$YEAR <- as.factor(avg.exp.county.both$YEAR)


ggplot(avg.exp.county.both, aes(x=YEAR, average.county.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Exposure Index")+
  ggtitle("Average Exposure Index by Dismissal Status on County
          Level Including Publics and Privates")



################################
# Exposure Index: District
################################

df.schools$pct.to.sum <- (df.schools$WHITE/df.schools$NUMSTUDS)*df.schools$BLACK

exp.district.level <- df.schools %>% 
  group_by(DISTRICT, YEAR, DISMISSED) %>%
  summarise(district.index=sum((WHITE/NUMSTUDS)*BLACK),
            black.district.total=sum(BLACK))
exp.district.level[is.na(exp.district.level)] <- 0
exp.district.level$district.index <- exp.district.level$district.index/exp.district.level$black.district.total
exp.district.level[is.na(exp.district.level)] <- 0


avg.exp.district.both <- exp.district.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.district.index = mean(district.index))
avg.exp.district.both$DISMISSED <- as.factor(avg.exp.district.both$DISMISSED)
avg.exp.district.both$YEAR <- as.factor(avg.exp.district.both$YEAR)


ggplot(avg.exp.district.both, aes(x=YEAR, average.district.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Exposure Index")+
  ggtitle("Average Exposure Index by Dismissal Status on District
          Level Including Publics and Privates")
```

# Publics Only

## Dissimilarity

```{r pub only: dissimilarity, echo=FALSE}

################################
# Dissimilarity Index: County
################################

pub.index <- pub.index %>% 
  group_by(COUNTY, YEAR) %>%
  mutate(county.total=sum(NUMSTUDS),
         black.county.total=sum(BLACK),
         white.county.total=sum(WHITE))

pub.index$pct.black.county.lvl <- pub.index$BLACK/pub.index$black.county.total
pub.index$pct.white.county.lvl <- pub.index$WHITE/pub.index$white.county.total
pub.index$pct.diff.county.lvl <- abs(pub.index$pct.black.county.lvl-pub.index$pct.white.county.lvl)
pub.index[is.na(pub.index)] <- 0

dissimilarity.county.level <- pub.index %>% 
  group_by(COUNTY, YEAR, DISMISSED) %>%
  summarise(county.index=0.5*sum(pct.diff.county.lvl))

avg.diss.county.both <- dissimilarity.county.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.county.index = mean(county.index))
avg.diss.county.both$DISMISSED <- as.factor(avg.diss.county.both$DISMISSED)
avg.diss.county.both$YEAR <- as.factor(avg.diss.county.both$YEAR)


ggplot(avg.diss.county.both, aes(x=YEAR, average.county.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Dissimilarity Index")+
  ggtitle("Average Dissimilarity Index by Dismissal Status on County
          Level Including Publics Only")


################################
# Dissimilarity Index: District
################################

pub.index <- pub.index %>% 
  group_by(DISTRICT, YEAR) %>%
  mutate(district.total=sum(NUMSTUDS),
         black.district.total=sum(BLACK),
         white.district.total=sum(WHITE))

pub.index$pct.black.district.lvl <- pub.index$BLACK/pub.index$black.district.total
pub.index$pct.white.district.lvl <- pub.index$WHITE/pub.index$white.district.total
pub.index$pct.diff.district.lvl <- abs(pub.index$pct.black.district.lvl-pub.index$pct.white.district.lvl)
pub.index[is.na(pub.index)] <- 0

dissimilarity.district.level <- pub.index %>% 
  group_by(DISTRICT, YEAR, DISMISSED) %>%
  summarise(district.index=0.5*sum(pct.diff.district.lvl))

avg.diss.district.both <- dissimilarity.district.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.district.index = mean(district.index))
avg.diss.district.both$DISMISSED <- as.factor(avg.diss.district.both$DISMISSED)
avg.diss.district.both$YEAR <- as.factor(avg.diss.district.both$YEAR)

ggplot(avg.diss.district.both, aes(x=YEAR, average.district.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Dissimilarity Index")+
  ggtitle("Average Dissimilarity Index by Dismissal Status on District
          Level Including Publics Only")

```

## Exposure
```{r pub only: exposure, echo=FALSE}

################################
# Exposure Index
################################

pub.index$pct.to.sum <- (pub.index$WHITE/pub.index$NUMSTUDS)*pub.index$BLACK

exp.county.level <- pub.index %>% 
  group_by(COUNTY, YEAR, DISMISSED) %>%
  summarise(county.index=sum((WHITE/NUMSTUDS)*BLACK),
            black.county.total=sum(BLACK))
exp.county.level[is.na(exp.county.level)] <- 0
exp.county.level$county.index <- exp.county.level$county.index/exp.county.level$black.county.total
exp.county.level[is.na(exp.county.level)] <- 0


avg.exp.county.both <- exp.county.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.county.index = mean(county.index))
avg.exp.county.both$DISMISSED <- as.factor(avg.exp.county.both$DISMISSED)
avg.exp.county.both$YEAR <- as.factor(avg.exp.county.both$YEAR)


ggplot(avg.exp.county.both, aes(x=YEAR, average.county.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Exposure Index")+
  ggtitle("Average Exposure Index by Dismissal Status on County
          Level Including Publics Only")

################################
# Exposure Index
################################

pub.index$pct.to.sum <- (pub.index$WHITE/pub.index$NUMSTUDS)*pub.index$BLACK

exp.district.level <- pub.index %>% 
  group_by(DISTRICT, YEAR, DISMISSED) %>%
  summarise(district.index=sum((WHITE/NUMSTUDS)*BLACK),
            black.district.total=sum(BLACK))
exp.district.level[is.na(exp.district.level)] <- 0
exp.district.level$district.index <- exp.district.level$district.index/exp.district.level$black.district.total
exp.district.level[is.na(exp.district.level)] <- 0


avg.exp.district.both <- exp.district.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.district.index = mean(district.index))
avg.exp.district.both$DISMISSED <- as.factor(avg.exp.district.both$DISMISSED)
avg.exp.district.both$YEAR <- as.factor(avg.exp.district.both$YEAR)


ggplot(avg.exp.district.both, aes(x=YEAR, average.district.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Exposure Index")+
  ggtitle("Average Exposure Index by Dismissal Status on District
          Level Including Publics Only")
```


# Privates Only

## Dissimilarity

```{r pvt only: dissimilarity, echo=FALSE}

################################
# Dissimilarity Index: County
################################

pvt.index <- pvt.index %>% 
  group_by(COUNTY, YEAR) %>%
  mutate(county.total=sum(NUMSTUDS),
         black.county.total=sum(BLACK),
         white.county.total=sum(WHITE))

pvt.index$pct.black.county.lvl <- pvt.index$BLACK/pvt.index$black.county.total
pvt.index$pct.white.county.lvl <- pvt.index$WHITE/pvt.index$white.county.total
pvt.index$pct.diff.county.lvl <- abs(pvt.index$pct.black.county.lvl-pvt.index$pct.white.county.lvl)
pvt.index[is.na(pvt.index)] <- 0

dissimilarity.county.level <- pvt.index %>% 
  group_by(COUNTY, YEAR, DISMISSED) %>%
  summarise(county.index=0.5*sum(pct.diff.county.lvl))

avg.diss.county.both <- dissimilarity.county.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.county.index = mean(county.index))
avg.diss.county.both$DISMISSED <- as.factor(avg.diss.county.both$DISMISSED)
avg.diss.county.both$YEAR <- as.factor(avg.diss.county.both$YEAR)


ggplot(avg.diss.county.both, aes(x=YEAR, average.county.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Dissimilarity Index")+
  ggtitle("Average Dissimilarity Index by Dismissal Status on County
          Level Including Privates Only")

################################
# Dissimilarity Index: District
################################

pvt.index <- pvt.index %>% 
  group_by(DISTRICT, YEAR) %>%
  mutate(district.total=sum(NUMSTUDS),
         black.district.total=sum(BLACK),
         white.district.total=sum(WHITE))

pvt.index$pct.black.district.lvl <- pvt.index$BLACK/pvt.index$black.district.total
pvt.index$pct.white.district.lvl <- pvt.index$WHITE/pvt.index$white.district.total
pvt.index$pct.diff.district.lvl <- abs(pvt.index$pct.black.district.lvl-pvt.index$pct.white.district.lvl)
pvt.index[is.na(pvt.index)] <- 0

dissimilarity.district.level <- pvt.index %>% 
  group_by(DISTRICT, YEAR, DISMISSED) %>%
  summarise(district.index=0.5*sum(pct.diff.district.lvl))

avg.diss.district.both <- dissimilarity.district.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.district.index = mean(district.index))
avg.diss.district.both$DISMISSED <- as.factor(avg.diss.district.both$DISMISSED)
avg.diss.district.both$YEAR <- as.factor(avg.diss.district.both$YEAR)

ggplot(avg.diss.district.both, aes(x=YEAR, average.district.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Dissimilarity Index")+
  ggtitle("Average Dissimilarity Index by Dismissal Status on District
          Level Including Privates Only")


```

## Exposure
```{r pvt only: exposure, echo=FALSE}

################################
# Exposure Index: County
################################

pvt.index$pct.to.sum <- (pvt.index$WHITE/pvt.index$NUMSTUDS)*pvt.index$BLACK

exp.county.level <- pvt.index %>% 
  group_by(COUNTY, YEAR, DISMISSED) %>%
  summarise(county.index=sum((WHITE/NUMSTUDS)*BLACK),
            black.county.total=sum(BLACK))
exp.county.level[is.na(exp.county.level)] <- 0
exp.county.level$county.index <- exp.county.level$county.index/exp.county.level$black.county.total
exp.county.level[is.na(exp.county.level)] <- 0


avg.exp.county.both <- exp.county.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.county.index = mean(county.index))
avg.exp.county.both$DISMISSED <- as.factor(avg.exp.county.both$DISMISSED)
avg.exp.county.both$YEAR <- as.factor(avg.exp.county.both$YEAR)


ggplot(avg.exp.county.both, aes(x=YEAR, average.county.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Exposure Index")+
  ggtitle("Average Exposure Index by Dismissal Status on County
          Level Including Privates Only")

################################
# Exposure Index: District
################################

pvt.index$pct.to.sum <- (pvt.index$WHITE/pvt.index$NUMSTUDS)*pvt.index$BLACK

exp.district.level <- pvt.index %>% 
  group_by(DISTRICT, YEAR, DISMISSED) %>%
  summarise(district.index=sum((WHITE/NUMSTUDS)*BLACK),
            black.district.total=sum(BLACK))
exp.district.level[is.na(exp.district.level)] <- 0
exp.district.level$district.index <- exp.district.level$district.index/exp.district.level$black.district.total
exp.district.level[is.na(exp.district.level)] <- 0


avg.exp.district.both <- exp.district.level %>%
  group_by(YEAR, DISMISSED) %>%
  summarise(average.district.index = mean(district.index))
avg.exp.district.both$DISMISSED <- as.factor(avg.exp.district.both$DISMISSED)
avg.exp.district.both$YEAR <- as.factor(avg.exp.district.both$YEAR)


ggplot(avg.exp.district.both, aes(x=YEAR, average.district.index, group=DISMISSED, color=DISMISSED)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept=5)+
  xlab("Year") +
  ylab("Average Exposure Index")+
  ggtitle("Average Exposure Index by Dismissal Status on District
          Level Including Privates Only")
```