---
title: "private_schools"
author: "Mike Finnegan"
date: "11/20/2017"
output: html_document
---

## Load Data into R
``` {r import data}
invol_deseg <- data.table::fread('~/Dropbox/PkgData/SchoolData/desegregation-orders/invol_data.csv')
vol_deseg <- data.table::fread('~/Dropbox/PkgData/SchoolData/desegregation-orders/vol_data.csv')
invol_deseg<- invol_deseg[which(invol_deseg$State=='AL')]

setwd("~/Documents/Github/private")
export(characterize(import("pss8990_pu.sas7bdat")), "pss89.csv")
export(characterize(import("pss9192_pu.sas7bdat")), "pss91.csv")
export(characterize(import("pss9394_pu.sas7bdat")), "pss93.csv")
export(characterize(import("pss9596_pu.sas7bdat")), "pss95.csv")
export(characterize(import("pss9798_pu.sas7bdat")), "pss97.csv")
export(characterize(import("pss9900_pu.sas7bdat")), "pss99.csv")
export(characterize(import("pss0102_pu.sas7bdat")), "pss01.csv")
export(characterize(import("pss0304_pu.sas7bdat")), "pss03.csv")
export(characterize(import("pss0506_pu.sas7bdat")), "pss05.csv")
export(characterize(import("pss0708_pu.sas7bdat")), "pss07.csv")
export(characterize(import("pss0910_pu.sas7bdat")), "pss09.csv")
export(characterize(import("pss1112_pu.sas7bdat")), "pss11.csv")

pvt_89 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss89.csv")
pvt_91 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss91.csv")
pvt_93 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss93.csv")
pvt_95 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss95.csv")
pvt_97 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss97.csv")
pvt_99 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss99.csv")
pvt_01 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss01.csv")
pvt_03 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss03.csv")
pvt_05 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss05.csv")
pvt_07 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss07.csv")
pvt_09 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss09.csv")
pvt_11 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss11.csv")
pvt_13 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss13.csv")
pvt_15 <- read.csv("~/Dropbox/PkgData/SchoolData/private/pss15.csv")
```

## Reduce Size of Data
```{r filter data}
#Only select private schools in Alabama 
pvt_89 <- filter(pvt_89, pvt_89$STATE=='AL')
pvt_91 <- filter(pvt_91, pvt_91$ESTABB=='AL')
pvt_93 <- filter(pvt_93, pvt_93$PSTABB=='AL')
pvt_95 <- filter(pvt_95, pvt_95$QSTABB=='AL')
pvt_97 <- filter(pvt_97, pvt_97$RSTABB=='AL')
pvt_99 <- filter(pvt_99, pvt_99$SSTABB=='AL')
pvt_01 <- filter(pvt_01, pvt_01$PSTABB=='AL')
pvt_03 <- filter(pvt_03, pvt_03$PSTABB=='AL')
pvt_05 <- filter(pvt_05, pvt_05$pstabb=='AL')
pvt_07 <- filter(pvt_07, pvt_07$pstabb=='AL')
pvt_09 <- filter(pvt_09, pvt_09$pstabb=='AL')
pvt_11 <- filter(pvt_11, pvt_11$pstabb=='AL')
pvt_13 <- filter(pvt_13, pvt_13$PSTABB=='AL')
pvt_15 <- filter(pvt_15, pvt_15$pstabb=='AL')

#select first 5 digits of zipcode
pvt_89$EZIP <- substr(pvt_89$EZIP, 1, 5)
pvt_91$EZIP <- substr(pvt_91$EZIP, 1, 5)
pvt_93$PZIP <- substr(pvt_93$PZIP, 1, 5)
pvt_95$QZIP <- substr(pvt_99$QZIP, 1, 5)
pvt_97$RZIP <- substr(pvt_97$RZIP, 1, 5)
pvt_99$SZIP <- substr(pvt_99$SZIP, 1, 5)
pvt_01$PZIP <- substr(pvt_01$PZIP, 1, 5)
pvt_03$PZIP <- substr(pvt_03$PZIP, 1, 5)
pvt_05$pzip <- substr(pvt_05$pzip, 1, 5)
pvt_07$pzip <- substr(pvt_07$pzip, 1, 5)
pvt_09$pzip <- substr(pvt_09$pzip, 1, 5)
pvt_11$pzip <- substr(pvt_11$pzip, 1, 5)
pvt_13$PZIP <- substr(pvt_13$PZIP, 1, 5)
pvt_15$pzip <- substr(pvt_15$pzip, 1, 5)

## consolidate variables
keep89 <- c("EPIN", "CITY", "STATE", "EZIP", "PSS064",
            "NUMSTUDS", "STTCH_RT", "HIGRADE") 
pvt_89 <- pvt_89[keep89]
names(pvt_89)[names(pvt_89)=="PSS064"] <- "religious"
names(pvt_89)[names(pvt_89)=="EPIN"] <- "pin"
names(pvt_89)[names(pvt_89)=="STATE"] <- "STABB"
pvt_89$INST <- NA
pvt_89$ADDRS <- NA
pvt_89$CNT <- NA
pvt_89$MINOR <- NA
pvt_89$zip89 <- substr(pvt_89$zip89, 1, 5)

keep91 <- c("EPIN", "ECITY", "ESTABB", "ECNTNM", "EZIP", "PSC063",
            "NUMSTUDS", "STTCH_RT", "HIGRADE") 
pvt_91 <- pvt_91[keep91]
names(pvt_91)[names(pvt_91)=="PSC063"] <- "religious"
names(pvt_91)[names(pvt_91)=="EPIN"] <- "pin"
pvt_91$INST <- NA
pvt_91$ADDRS <- NA
pvt_91$MINOR <- NA
pvt_91$zip91 <- substr(pvt_91$zip91, 1, 5)

keep93 <- c("PPIN", "PCITY", "PSTABB", "PCNTNM", "PZIP", "P380",
            "NUMSTUDS", "STTCH_RT", "HIGRADE",  "P_INDIAN", "P_ASIAN", "P_HISP", "P_BLACK", "P_WHITE",
            "PMINOR") 
pvt_93 <- pvt_93[keep93]
names(pvt_93)[names(pvt_93)=="P380"] <- "religious"
names(pvt_93)[names(pvt_93)=="PPIN"] <- "pin"
pvt_93$INST <- NA
pvt_93$ADDRS <- NA
pvt_93$zip93 <- substr(pvt_93$zip93, 1, 5)

keep95 <- c("QPIN", "QCITY", "QSTABB", "QCNTNM", "QZIP", "P430",
            "NUMSTUDS", "STTCH_RT",  "P_INDIAN", "P_ASIAN", "P_HISP",
            "P_BLACK", "P_WHITE", "PERMINOR", "QHIGR6") 
pvt_95 <- pvt_95[keep95]
names(pvt_95)[names(pvt_95)=="P430"] <- "religious"
names(pvt_95)[names(pvt_95)=="QPIN"] <- "pin"
names(pvt_95)[names(pvt_95)=="QHIGR6"] <- "HIGR1996"
pvt_95$INST <- NA
pvt_95$ADDRS <- NA
pvt_95$zip95 <- substr(pvt_95$zip95, 1, 5)

keep97 <- c("RPIN", "RCITY", "RSTABB", "RCNTNM", "RZIP", "P340", "P430",
            "NUMSTUDS", "STTCH_RT",  "P_INDIAN", "P_ASIAN", "P_HISP",
            "P_BLACK", "P_WHITE", "PMINOR", "HIGR8") 
pvt_97 <- pvt_97[keep97]
names(pvt_97)[names(pvt_97)=="P340"] <- "MALES"
names(pvt_97)[names(pvt_97)=="P430"] <- "religious"
names(pvt_97)[names(pvt_97)=="RPIN"] <- "pin"
names(pvt_97)[names(pvt_97)=="HIGR8"] <- "HIGR1998"
pvt_97$INST <- NA
pvt_97$ADDRS <- NA
pvt_97$zip97 <- substr(pvt_97$zip97, 1, 5)

keep99 <- c("SPIN", "SCITY", "SSTABB", "SCNTYNM", "SZIP", "P340", "P360", 
            "P361", "P430", "NUMSTUDS", "STTCH_RT",  "P_INDIAN", "P_ASIAN", 
            "P_HISP", "P_BLACK", "P_WHITE", "PMINOR", "HIGR2000") 
pvt_99 <- pvt_99[keep99]
names(pvt_99)[names(pvt_99)=="P340"] <- "MALES"
names(pvt_99)[names(pvt_99)=="P360"] <- "pct_4year"
names(pvt_99)[names(pvt_99)=="P361"] <- "pct_2year"
names(pvt_99)[names(pvt_99)=="P430"] <- "religious"
names(pvt_99)[names(pvt_99)=="SZIP"] <- "zip99"
pvt_99$INST <- NA
pvt_99$ADDRS <- NA
pvt_99$zip99 <- substr(pvt_99$zip99, 1, 5)

keep01 <- c("PPIN", "PCITY", "PSTABB", "PCNTNM", "PZIP", "P340", "P360", 
            "P361", "P430", "NUMSTUDS", "STTCH_RT",  "P_INDIAN", "P_ASIAN", 
            "P_HISP", "P_BLACK", "P_WHITE", "PMINOR", "HIGR2002") 
pvt_01 <- pvt_01[keep01]
names(pvt_01)[names(pvt_01)=="P340"] <- "MALES"
names(pvt_01)[names(pvt_01)=="P360"] <- "pct_4year"
names(pvt_01)[names(pvt_01)=="P361"] <- "pct_2year"
names(pvt_01)[names(pvt_01)=="P430"] <- "religious"
names(pvt_01)[names(pvt_01)=="PZIP"] <- "zip01"
pvt_01$INST <- NA
pvt_01$ADDRS <- NA
pvt_01$zip01 <- substr(pvt_01$zip01, 1, 5)

keep03 <- c("PPIN", "PCITY", "PSTABB", "PCNTNM",
            "PZIP", "P360", "P361", "P430", "MALES", "NUMSTUDS",
            "P_INDIAN", "P_ASIAN", "P_HISP", "P_BLACK", "P_WHITE",
            "PMINOR", "STTCH_RT", "HIGR2004") 
pvt_03 <- pvt_03[keep03]
names(pvt_03)[names(pvt_03)=="P360"] <- "pct_4year"
names(pvt_03)[names(pvt_03)=="P361"] <- "pct_2year"
names(pvt_03)[names(pvt_03)=="P430"] <- "religious"
names(pvt_03)[names(pvt_03)=="PPIN"] <- "pin"
pvt_03$INST <- NA
pvt_03$ADDRS <- NA
pvt_03$zip03 <- substr(pvt_03$zip03, 1, 5)

keep05 <- c("ppin", "pinst", "pcity", "pstabb", "pcntnm",
            "pzip","latitude", "longitude", "P360", "P361", "P430",
            "males", "numstuds", "p_indian", "p_asian", "p_hisp", 
            "p_black", "p_white", "pminor", "sttch_rt", "higr2006")
pvt_05 <- pvt_05[keep05]
names(pvt_05)[names(pvt_05)=="P360"] <- "pct_4year"
names(pvt_05)[names(pvt_05)=="P361"] <- "pct_2year"
names(pvt_05)[names(pvt_05)=="P430"] <- "religious"
names(pvt_05)[names(pvt_05)=="ppin"] <- "pin"
pvt_05$ADDRS <- NA
pvt_05$zip05 <- substr(pvt_05$zip05, 1, 5)

keep07 <- c("ppin", "pinst", "pcity", "pstabb", "pcntnm",
            "pzip", "paddrs","latitude", "longitude", "P360", "P430",
            "males", "numstuds", "p_indian", "p_asian", "p_hisp", 
            "p_black", "p_white", "pminor", "sttch_rt", "higr2008")
pvt_07 <- pvt_07[keep07]
names(pvt_07)[names(pvt_07)=="P360"] <- "pct_4year"
names(pvt_07)[names(pvt_07)=="P430"] <- "religious"
names(pvt_07)[names(pvt_07)=="ppin"] <- "pin"
pvt_07$zip07 <- substr(pvt_07$zip07, 1, 5)

keep09 <- c("ppin", "pinst", "pcity", "pstabb", "pcntnm",
            "pzip", "paddrs", "latitude10", "longitude10", "P360", "P430",
            "males", "numstuds", "p_indian", "p_asian", "p_hisp", 
            "p_pacific","p_black", "p_white", "p_two", "sttch_rt", 
            "higr2010")
pvt_09 <- pvt_09[keep09]
names(pvt_09)[names(pvt_09)=="P360"] <- "pct_4year"
names(pvt_09)[names(pvt_09)=="P430"] <- "religious"
names(pvt_09)[names(pvt_09)=="ppin"] <- "pin"
names(pvt_09)[names(pvt_09)=="P_TWO"] <- "P_TR"
pvt_09$zip09 <- substr(pvt_09$zip09, 1, 5)

keep11 <- c("ppin", "pinst", "pcity", "pstabb", "pcntnm",
            "pzip", "paddrs", "latitude12", "longitude12", "p360", "p430",
            "males", "numstuds", "p_indian", "p_asian", "p_hisp", 
            "p_pacific","p_black", "p_white", "p_tr", "sttch_rt", 
            "higr2012")
pvt_11 <- pvt_11[keep11]
names(pvt_11)[names(pvt_11)=="p360"] <- "pct_4year"
names(pvt_11)[names(pvt_11)=="p430"] <- "religious"
names(pvt_11)[names(pvt_11)=="ppin"] <- "pin"
pvt_11$zip11 <- substr(pvt_11$zip11, 1, 5)

keep13 <- c("PPIN", "PINST", "PCITY", "PSTABB", "PCNTNM",
            "PZIP", "PADDRS", "LATITUDE14", "LONGITUDE14", "P360", "P430",
            "MALES", "NUMSTUDS", "P_INDIAN", "P_ASIAN", "P_PACIFIC",
            "P_HISP","P_BLACK", "P_WHITE", "P_TR", "STTCH_RT", 
            "HIGR2014")
pvt_13 <- pvt_13[keep13]
names(pvt_13)[names(pvt_13)=="P360"] <- "pct_4year"
names(pvt_13)[names(pvt_13)=="P430"] <- "religious"
names(pvt_13)[names(pvt_13)=="PPIN"] <- "pin"
pvt_13$zip13 <- substr(pvt_13$zip13, 1, 5)

keep15 <- c("ppin", "pinst", "pcity", "pstabb", "pcntnm",
            "pzip", "paddrs", "latitude16", "longitude16", "p360", "p430",
            "males", "numstuds", "p_indian", "p_asian", "p_hisp", 
            "p_pacific","p_black", "p_white", "p_tr", "sttch_rt", 
            "higr2016")
pvt_15 <- pvt_15[keep15]
names(pvt_15)[names(pvt_15)=="p360"] <- "pct_4year"
names(pvt_15)[names(pvt_15)=="p430"] <- "religious"
names(pvt_15)[names(pvt_15)=="ppin"] <- "pin"
pvt_15$zip15 <- substr(pvt_15$zip15, 1, 5)


pvt_89$YR <- '89'
pvt_91$YR <- '91'
pvt_93$YR <- '93'
pvt_95$YR <- '95'
pvt_97$YR <- '97'
pvt_99$YR <- '99'
pvt_01$YR <- '01'
pvt_03$YR <- '03'
pvt_05$YR <- '05'
pvt_07$YR <- '07'
pvt_09$YR <- '09'
pvt_11$YR <- '11'
pvt_13$YR <- '13'
pvt_15$YR <- '15'
```

## Standardize private school variables
```{r standardize}
l.pvt <- list(pvt_89, pvt_91, pvt_93, pvt_95, pvt_97, pvt_99, pvt_01, pvt_03, pvt_05, pvt_07, pvt_09, pvt_11, pvt_13, pvt_15)


funStandardizeFields <- function(dt){
  data.table::setnames(dt, names(dt), stringr::str_to_upper(names(dt)))
  data.table::setnames(dt, names(dt)[grep('HIGR', names(dt))], 'HIGR')
  data.table::setnames(dt, names(dt)[grep('ZIP', names(dt))], 'ZIP')
  data.table::setnames(dt, names(dt)[grep('CITY', names(dt))], 'CITY')
  if (length(grep('CNT', names(dt)))>0){
    data.table::setnames(dt, names(dt)[grep('CNT', names(dt))], 'COUNTY')
  }
  if (length(grep('STABB', names(dt)))>0){
    data.table::setnames(dt, names(dt)[grep('STABB', names(dt))], 'STATE')
  }
  if (length(grep('INST', names(dt)))>0){
    data.table::setnames(dt, names(dt)[grep('INST', names(dt))], 'SCHOOL')
  }
  if (length(grep('ADDRS', names(dt)))>0){
    data.table::setnames(dt, names(dt)[grep('ADDRS', names(dt))], 'ADDRESS')
  }
  if (length(grep('MINOR', names(dt)))>0){
    data.table::setnames(dt, names(dt)[grep('MINOR', names(dt))], 'P_MINOR')
  }
  if (length(grep('LATITUDE', names(dt)))>0){
    data.table::setnames(dt, names(dt)[grep('LATITUDE', names(dt))], 'LATITUDE')
    data.table::setnames(dt, names(dt)[grep('LONGITUDE', names(dt))], 'LONGITUDE')
  }
  dt$PIN <- as.character(dt$PIN)
  return(dt)
} 
l.pvt <- lapply(l.pvt, funStandardizeFields)
dt.pvt <- data.table::rbindlist(l.pvt, fill=TRUE, use.names=TRUE)
dt.pvt$STATE <- "AL"
#Recode non-parochial schools as having value of 0
dt.pvt$RELIGIOUS[dt.pvt$RELIGIOUS==2] <- 0
```

To assign latitude and longitude coordinates to schools that do not have any lat/long data to utilize, I have taken a data file containing the lat/long of the centroid of all U.S. zip codes. This data is from the 2010 U.S. census and the file was downloaded from http://www.nber.org/data/zip-code-distance-database.html.

## Assign Missing School Locations 
```{r geospatial corrections}

# To build zip 'best guess', I grab zipcode centroids data from the 2010 census
centroids <- read.csv("~/Box Sync/zip_centroids_2010census.csv")

funAssignGeo <- function(x.pin, dt.pvt, zip.locs){
  dt.pin <- dt.pvt[PIN==x.pin]
  
  # Check to see if missed survey for a given year
  n.years <- length(unique(dt.pvt$YR))
  n.pin.years <- nrow(dt.pin)
  if (n.years != n.pin.years){
    #unsure what to do next
  }

  # If unique zipcodes then try to assign lat long
  zips <- unique(dt.pin$ZIP)
  zips <- zips[!is.na(zips)]
  if (!(is.na(zips)) && length(zips==1)){
    lat.val <- unique(round(dt.pin$LATITUDE,digits = 3))
    long.val <- unique(round(dt.pin$LONGITUDE,digits = 3))
    if (length(lat.val[!is.na(lat.val)])>1){ # One lat long pair
      cat('non unique lat vals for', pin)
      # unsure what to do next
    } 
    if (length(lat.val[!is.na(lat.val)])==0){ # No lat long pair, so use zip centroid
      
      #Two line below running without warning, but they assign lat/long centroids to years that already have lat/long values, which is opposite of what is intended.
      
          #dt.pin$LATITUDE[dt.pin$PIN %in%centroids$zcta5] <- centroids$intptlat
          #dt.pin$LONGITUDE[dt.pin$PIN %in% centroids$zcta5] <- centroids$intptlon
        }
    }
  # print(head(dt.pin))
  return(dt.pin)
  }
}

PINS <- unique(dt.pvt$PIN)

l.clean <- lapply(PINS, function(x.pin) funAssignGeo(x.pin, dt.pvt, zip.locs))
dt.clean <- data.table::rbindlist(l.clean,use.names = TRUE, fill=TRUE)
```

## Fill Missing Values for School and County
```{fill School and County NA's}
dt.names <- na.omit(distinct(select(dt.clean, PIN, SCHOOL)))
for(id in 1:nrow(dt.names)){
  dt.clean$SCHOOL[dt.clean$PIN %in% dt.names$PIN[id]] <- dt.names$SCHOOL[id]
}

dt.cnty <- na.omit(distinct(select(dt.clean, PIN, COUNTY)))
for(id in 1:nrow(dt.cnty)){
  dt.clean$COUNTY[dt.clean$PIN %in% dt.cnty$PIN[id]] <- dt.cnty$COUNTY[id]
}
```

## Initial Graphs and Summary Statistics
```{r summary stats}
count.schools <- data.frame(table(dt.clean$YR))
count.schools$year <- c(2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 1989, 1991, 1993, 1995, 1997, 1999)

ggplot(count.schools, aes(x=year, Freq)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("# of Private Schools") +
  ggtitle("All Private Schools in Alabama from 1989 to 2015")

library(purrr)
summary.stats <- dt.clean %>% split(.$YR) %>% map(summary)
#Summary statistics stored in a list, in which each year is a member containing summary stats on all variabes grouped by that year.

#The summary stats below are similar, but more expansive, and make it easier to look only at one particular variable of interest (in this case the mean percent of black students in private schools) for each year.

library(psych)
summary.stats.pblack <- na.omit(describe.by(dt.clean$P_BLACK, group=dt.clean$YR, mat=TRUE))
summary.stats.pblack$year <- c(2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 1993, 1995, 1997, 1999)

ggplot(summary.stats.pblack, aes(x=year, mean)) +
  geom_line() +
  geom_point() +
  xlab("Year") +
  ylab("Mean Percent of Black Students") +
  ggtitle("Percent Black in Alabama Privates from 1989 to 2015")
```

